---
title: "Compare ENC between binary and weighted interactions"
output: html_notebook
author: AFC
---


**Can we approximate net effects of interactions from binary matrices?**


```{r}
source("simulate_dynamics.R")
```



## A simple case example

A weighted matrix and its net effects:

```{r}
M <- matrix(c(0,-2,4,1,
              3,0,0,2,
              -5,0,0,0,
              -0.5,-1,0,0),
            nrow = 4, byrow = TRUE)
Minv <- solve(M) # net effects from weighted matrix


M
Minv
```

The net effects of its binary matrix seem to approximate well its net effects:

```{r}
Msign <- sign(M) # binary matrix, retaining the sign
MsignInv <- solve(Msign) # net effects from binary matrix retaining the sign

MsignInv
```
Which is not the case if we dont keep the sign:

```{r}
Mpos <- abs(Msign) # positive binary matrix
MposInv <- solve(Mpos) # net effects from positive binary matrix

MposInv
```

## Inspect how well the net effects from weighted matrix are approximated by its binary versions



### Create scenarios of varying species richness (from 10 to 100, by 5):


```{r}

nsp = seq(from = 10, to = 100, by = 5)
scenario = seq(1:length(nsp))
replicates = c(1:5)

df = data.frame(scenario, nsp)

# Create the dataframe
df <- data.frame(scenario, nsp)

# Replicate each scenario and nsp values
df <- df[rep(1:nrow(df), each = length(replicates)), ]

rownames(df) <- NULL

df

```


### Loop to compute net effects for both weighted and binary matrices for each scenario (takes 1-2min):


```{r}


# Storing objects

w_mat <- vector("list", nrow(df))
w_net_mat <- vector("list", nrow(df))
b_mat <- vector("list", nrow(df))
b_net_mat <- vector("list", nrow(df))
plot <- vector("list", nrow(df))
correlation <- numeric(nrow(df))
signif <- character(nrow(df))


for (i in 1:nrow(df)) {
  
  print(paste(sprintf("iteration %s", i), "of", nrow(df)))
  
  nsp = df[i, "nsp"]
  
  
  weighted_mat <- sim_quantitative_network("predator-prey", nsp, 0.2, aij_params = c(0, 0.1), efficiency = 0.5, rho = 0)
  
  w_mat[[i]] = weighted_to_binary_neteffects(weighted_mat)$weighted
  w_net_mat[[i]] = weighted_to_binary_neteffects(weighted_mat)$net_effects_weighted
  b_mat[[i]] = weighted_to_binary_neteffects(weighted_mat)$binary
  b_net_mat[[i]] = weighted_to_binary_neteffects(weighted_mat)$net_efects_binary
  
  
  
  
  
  # Create plots
  
  # Convert matrix to dataframe
  w_net_df <- as.data.frame(w_net_mat[[i]])
  w_net_df$row <- rownames(w_net_df)
  
  b_net_df <- as.data.frame(b_net_mat[[i]])
  b_net_df$row <- rownames(b_net_df)
  
  # Reshape dataframe for ggplot
  w_net_df_long <- tidyr::pivot_longer(w_net_df, cols = -row, names_to = "column", values_to = "value")
  b_net_df_long <- tidyr::pivot_longer(b_net_df, cols = -row, names_to = "column", values_to = "value")
  
  
  plot_w <- ggplot(w_net_df_long, aes(x = column, y = row, fill = value)) +
    geom_tile() +
    scale_fill_gradient(low = "white", high = "blue") +
    labs(title = paste(nsp, "spp")) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  plot_b <- ggplot(b_net_df_long, aes(x = column, y = row, fill = value)) +
    geom_tile() +
    scale_fill_gradient(low = "white", high = "blue") +
    labs(title = paste(nsp, "spp")) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  
  
  
  if(df[i, "nsp"] <= 30){  # save plots only until sizes = 40
    
    plot[[i]] = ggarrange(
      
      plot_w,
      plot_b,
      
      labels = LETTERS[1:2]
      
    )
    
    
  }
  
  
  
  
  # compute correlation
  
  correlation[i] <- compute_procrustes_corr(w_net_mat[[i]], b_net_mat[[i]])$correlation
  
  pvalue <- compute_procrustes_corr(w_net_mat[[i]], b_net_mat[[i]])$correlation
  
  if (pvalue <= 0.05) {
    signif[i] <- "sign"
  } else {
    signif[i] <- "non sign"
  }
  
}

```


> Note: I get the warning "Non euclidean distance" when computing the Procrustes



### Show plots comparing net effects' matrices computed from either weighted (A) or binary (B)

Number of species goes from 10 to 100, by 5. For each size, there are 3 different network comparisons.
Run the line below to plot any comparison. Positions in the list are: 1-10 (10spp), 11-20 (15spp), etc.

```{r}
plot[[2]]
```


### check correlations

```{r}

df_plot <- data.frame("nsp" = df$nsp, correlation, signif)

ggplot(df_plot, aes(x = nsp, y = correlation, color = signif)) +
  geom_point() +
  scale_color_manual(values = c("Significant" = "red", "Non-significant" = "black")) +
  labs(x = "Species richness", y = "Procrustes corr net eff (weighted vs binary)") +
  theme_classic()+
  ylim(0,1)




```

